---
title: "Using R and Python in One Document"
author: "Tahir Hussain"
date: 2026-09-28
---

Data: [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/), Palmer Station Antarctica LTER, released under the [CC0](https://creativecommons.org/publicdomain/zero/1.0/) licence. Cite as Horst AM, Hill AP, Gorman KB (2020). I use the copy that ships inside the `palmerpenguins` R package, so nothing is downloaded when the site is built.

In this post I use R and Python in the same document. R prepares a summary of the penguin data, Python reads that summary and works with it, and then R reads a result that Python computed. The two languages talk to each other through the `reticulate` package.

## Set up R and point it at the project's Python

```{r}
#| label: setup
#| message: false
#| warning: false
library(reticulate)
library(palmerpenguins)
library(dplyr)
library(ggplot2)
library(knitr)

# Use the Python from this project's .venv (the post runs from its own folder)
use_virtualenv("../../.venv", required = TRUE)
```

## Step 1: R computes a summary

```{r}
#| label: r-summary
species_summary <- penguins |>
  filter(!is.na(body_mass_g), !is.na(flipper_length_mm)) |>
  group_by(species) |>
  summarise(
    n = n(),
    mean_body_mass_g = mean(body_mass_g),
    mean_flipper_length_mm = mean(flipper_length_mm)
  ) |>
  as.data.frame()

species_summary
```

## Step 2: Python reads the R object

In Python, an R object is available as `r.name`. Here `r.species_summary` is the data frame from the chunk above, seen as a pandas DataFrame.

```{python}
#| label: python-work
import matplotlib.pyplot as plt

df = r.species_summary.copy() # pyrefly: ignore
df["species"] = df["species"].astype(str)
df["mass_per_mm_flipper"] = (
    df["mean_body_mass_g"] / df["mean_flipper_length_mm"]
).round(2)

# Objects created here can be read back in R with py$name
python_summary = df
heaviest = str(df.loc[df["mean_body_mass_g"].idxmax(), "species"])

df
```

```{python}
#| label: fig-python-bar
#| fig-cap: "Average body mass by species, drawn in Python from a summary computed in R."
fig, ax = plt.subplots(figsize=(6, 4))
ax.bar(df["species"], df["mean_body_mass_g"])
ax.set_xlabel("Species")
ax.set_ylabel("Mean body mass (g)")
plt.show()
```

## Step 3: R reads the Python result

In R, a Python object is available as `py$name`. The table below was built in Python, and the species named in the text comes from a Python variable.

```{r}
#| label: r-uses-python
py$python_summary |>
  kable(
    digits = 2,
    caption = "Summary table computed in Python and displayed by R."
  )
```

The heaviest species on average is **`r py$heaviest`**. That value was computed in Python and read here in R.

## What I found

Gentoo penguins are the heaviest species on average, and the bar chart and table agree. The main point of this post is the workflow: R did the data preparation, Python did the extra calculation and plotting, and R displayed the results, all in one document.
